\large{\bf Looking forwards and backwards through locally regulated populations }

### 2022-04-25 forwards-and-backwards.tex
tweaks

\comment{This corollary was the previous theorem in the previous draft.
    Besides this proposed set-up, we have the additional options of
    (a) renaming the corollary a theorem and only proving it,
    or (b) presenting only the corollary here (as a theorem)
    and presenting the more general theorem later on.
}

\begin{corollary} \label{cor:wavefront}
    In addition to the assumptions of Theorem~\ref{thm:lineages},
    suppose that the population process is described by a traveling wave with speed $\wavespeed$,
    i.e., the population has density
    $\varphi(t, x) = w(x - t \wavespeed)$
    where $w$ solves
    \begin{align*}
        r \Delta (\gamma w) + w F(w) + \wavespeed \cdot \grad w = 0 .
    \end{align*}
    Then the generator $\Lgen$
    of the motion of a linages in this population is
    \begin{align}
        \Lgen f
        &=
        r \gamma
        \left(
            \Delta f
            +
            2 \grad \log (\gamma w)
            \cdot \grad f
        \right)
        + \wavespeed \cdot \grad f .
    \end{align}
\end{corollary}

@@

In this section we prove Theorem~\ref{thm:lineages}.
The argument follows directly from the discussion in Section \ref{sec:limiting_lines_of_descent}
(and so maybe actually that content should be down here).

@@

@@

Writing $\hat{T}_{t,s}$ for the time-inhomogeneous semigroup
corresponding to the motion of ancestral lineages backwards in time
\comment{(Use notation for the motion of a lineage introduced above.)}
(that is, $\hat{T}_{t,s} f(y) = \int p(t,s,x,y) f(x) dy$),
we can write this as 
$$
    \int f(y)\varphi_t(y)\hat{T}_{t,s}g(y)dy
    =
    \int g(x) \varphi_s(x)
        \IE_{s,x} \left[
            \exp\left(
                \int_s^t(F+\gamma\DG r)(Y_u)du
            \right)f(Y_t)
        \right]dx.
$$
Differentiating with respect to $t$ at $t=s$, and writing 
$\hat{\mathcal L}_s$ for the generator of $\hat{T}_{t,s}$ at time $s$, we find
\begin{align*}
&
    \int f(y) \left\{
        \varphi_s(y) \widehat{\mathcal L}_s g(y)
        + \left(
            r(y) \DG^*(\gamma \varphi_s)(y) + \varphi_s(y) F(y)
        \right) g(y)
    \right\} dy \\
&\qquad =
    \int g(x) \varphi_s(x) \left(
        {\mathcal L}^Y_s f(x) + (F(x)+\gamma(x) \DG r(x)) f(x)
    \right) dx \\
&\qquad =
    \int f(x) \left(
        ({\mathcal L}^Y_s)^* (\varphi_s g)(x) + (F(x) + \gamma(x) \DG r(x)) \varphi_s(x) g(x)
    \right) dx ,
\end{align*}
where $({\mathcal L}^Y_s)^*$ is the adjoint of ${\mathcal L}^Y_s$
and we have used that $\varphi_s$ solves $\dot \varphi = r \DG^*(\gamma \varphi) + F \varphi$.
Since $f$ was arbitrary,
\begin{eqnarray*}
\hat{\mathcal L}_s g
    &=&
    \frac{1}{\varphi_s} \left[
        ({\mathcal L}^Y_s)^* (\varphi_s g)
        + \gamma \varphi_s g \DG(r)
        - r g \DG^*(\gamma \varphi_s)
    \right] .
\end{eqnarray*}
(Note that the $\phi_s F g$ terms have cancelled.)
Since the adjoint of ${\mathcal L}^Y_s$ is
\begin{align*}
    ({\mathcal L}^Y_s)^* f
    &=
    r \DG^* (\gamma f) - \gamma f \DG r ,
\end{align*}
we can rewrite the generator of a lineage as 
\begin{eqnarray*}
\hat{\mathcal L}_s g
    &=&
    \frac{r}{\varphi_s} \left[
        \DG^* (\gamma \varphi_s g)
        - g \DG^*(\gamma \varphi_s)
    \right] .
\end{eqnarray*}
This is equation \eqref{eqn:lineage_generator}.

To simplify to equation \eqref{eqn:lineage_generator2},
this, first define
$
    \DD f(x) = \sum_{ij} \covq_{ij} \partial_{ij} f(x),
$
% and note that $\DD$ satisfies the following identity:
% \begin{align}
%     \DD(fg)
%     &=
%     f \DD g
%     + 2 (\covq \grad f) \cdot \grad g
%     + g \DD f .
% \end{align}
and so the adjoint of $\DD$ is
$$
    \DD^* f(x)
    =
    \sum_{ij} \partial_{ij} (\covq_{ij} f(x)) .
$$
Note that $\DD^*$ satisfies the following identity:
\begin{align*}
    \DD^*(fg)
    &=
    \sum_{ij} \left\{
        g \partial_{ij} (\covq_{ij} f)
        + 2 \partial_{i} (\covq_{ij} f) \partial_j(g)
        + \covq_{ij} f \partial_{ij} g
    \right\} \\
    &=
    \sum_{ij} \left\{
        g \partial_{ij} (\covq_{ij} f)
        + 2 f \partial_{i} (\covq_{ij}) \partial_j(g)
        + 2 \covq_{ij} \partial_{i} (f) \partial_j(g)
        + \covq_{ij} f \partial_{ij} g
    \right\} \\
    &=
    g \DD^* f
    + 2 f \vec{c} \cdot \grad g
    + 2 (C \grad f) \cdot \grad g
    + f \DD g ,
\end{align*}
where $\vec{c}_j = \sum_i \partial_j C_{ij}$.
So, with $f = \gamma \varphi_s$,
\begin{eqnarray*}
\hat{\mathcal L}_s g
    &=&
    \frac{r}{\varphi_s} \left[
        \DD^*(\gamma \varphi_s g) - \grad \cdot (\gamma \varphi_s g \meanq)
        - g \DD^*(\gamma \varphi_s) + g \grad \cdot (\gamma \varphi_s \meanq)
    \right] \\
    &=&
    \frac{r}{\varphi_s} \left[
        \gamma \varphi_s \DD g
        + 2 \gamma \varphi_s \vec{c} \cdot \grad g
        + 2 (C \grad (\gamma \varphi_s)) \cdot \grad g
        - \gamma \varphi_s \meanq \cdot \grad g
    \right] \\
    &=&
    r \gamma \left[
        \DD g
        + 2 \vec{c} \cdot \grad g
        + 2 (C \grad \log(\gamma \varphi_s)) \cdot \grad g
        - \meanq \cdot \grad g
    \right] ,
\end{eqnarray*}
which is equation \eqref{eqn:lineage_generator}.

>>>>

\begin{corollary} \label{cor:wavefront}
    In addition to the assumptions of Corollary~\ref{cor:lineages_simple},
    suppose that the population process is described by a traveling wave with speed $\wavespeed$,
    i.e., the population has density
    $\varphi(t, x) = w(x - t \wavespeed)$
    where $w$ solves
    \begin{align*}
        r \Delta (\gamma w) + w F(w) + \wavespeed \cdot \grad w = 0 .
    \end{align*}
    Then the generator $\Lgen$
    of the motion of a linages in this population is
    \begin{align}
        \Lgen f
        &=
        r \gamma
        \left(
            \Delta f
            +
            2 \grad \log (\gamma w)
            \cdot \grad f
        \right)
        + \wavespeed \cdot \grad f .
    \end{align}
\end{corollary}

@@

In this section we prove Theorem~\ref{thm:lineages}.
The argument follows directly from the discussion in Section \ref{sec:limiting_lines_of_descent}.

@@

\comment{(TODO: Use notation for the motion of a lineage introduced in statement of theorem.)}

@@

Writing $\hat{T}_{t,s}$ for the time-inhomogeneous semigroup
corresponding to the motion of ancestral lineages backwards in time
(that is, $\hat{T}_{t,s} f(y) = \int p(t,s,x,y) f(x) dy$),
we can write this as 
$$
    \int f(y)\varphi_t(y)\hat{T}_{t,s}g(y)dy
    =
    \int g(x) \varphi_s(x)
        \IE_{s,x} \left[
            \exp\left(
                \int_s^t(F+\gamma\DG r)(Y_u)du
            \right)f(Y_t)
        \right]dx.
$$
Differentiating with respect to $t$ at $t=s$, and writing 
$\Lgen_s$ for the generator of $\hat{T}_{t,s}$ at time $s$, we find
\begin{align*}
&
    \int f(y) \left\{
        \varphi_s(y) \Lgen_s g(y)
        + \left(
            r(y) \DG^*(\gamma \varphi_s)(y) + \varphi_s(y) F(y)
        \right) g(y)
    \right\} dy \\
&\qquad =
    \int g(x) \varphi_s(x) \left(
        {\mathcal L}^Y_s f(x) + (F(x)+\gamma(x) \DG r(x)) f(x)
    \right) dx \\
&\qquad =
    \int f(x) \left(
        ({\mathcal L}^Y_s)^* (\varphi_s g)(x) + (F(x) + \gamma(x) \DG r(x)) \varphi_s(x) g(x)
    \right) dx ,
\end{align*}
where $({\mathcal L}^Y_s)^*$ is the adjoint of ${\mathcal L}^Y_s$
and we have used that $\varphi_s$ solves $\dot \varphi = r \DG^*(\gamma \varphi) + F \varphi$.
Since $f$ was arbitrary,
\begin{eqnarray*}
\Lgen_s g
    &=&
    \frac{1}{\varphi_s} \left[
        ({\mathcal L}^Y_s)^* (\varphi_s g)
        + \gamma \varphi_s g \DG(r)
        - r g \DG^*(\gamma \varphi_s)
    \right] .
\end{eqnarray*}
(Note that the $\phi_s F g$ terms have cancelled.)
Since the adjoint of ${\mathcal L}^Y_s$ is
\begin{align*}
    ({\mathcal L}^Y_s)^* f
    &=
    r \DG^* (\gamma f) - \gamma f \DG r ,
\end{align*}
we can rewrite the generator of a lineage as 
\begin{eqnarray*}
\Lgen_s g
    &=&
    \frac{r}{\varphi_s} \left[
        \DG^* (\gamma \varphi_s g)
        - g \DG^*(\gamma \varphi_s)
    \right] .
\end{eqnarray*}
This is equation \eqref{eqn:lineage_generator}.

To simplify to equation \eqref{eqn:lineage_generator2},
this, first define
$
    \DD f(x) = \sum_{ij} \covq_{ij} \partial_{ij} f(x),
$
% and note that $\DD$ satisfies the following identity:
% \begin{align}
%     \DD(fg)
%     &=
%     f \DD g
%     + 2 (\covq \grad f) \cdot \grad g
%     + g \DD f .
% \end{align}
and so the adjoint of $\DD$ is
$$
    \DD^* f(x)
    =
    \sum_{ij} \partial_{ij} (\covq_{ij} f(x)) .
$$
Note that $\DD^*$ satisfies the following identity:
\begin{align*}
    \DD^*(fg)
%     &=
%     \sum_{ij} \left\{
%         g \partial_{ij} (\covq_{ij} f)
%         + 2 \partial_{i} (\covq_{ij} f) \partial_j(g)
%         + \covq_{ij} f \partial_{ij} g
%     \right\} \\
    &=
    \sum_{ij} \left\{
        g \partial_{ij} (\covq_{ij} f)
        + 2 f \partial_{i} (\covq_{ij}) \partial_j(g)
        + 2 \covq_{ij} \partial_{i} (f) \partial_j(g)
        + \covq_{ij} f \partial_{ij} g
    \right\} \\
    &=
    g \DD^* f
    + 2 f \vec{c} \cdot \grad g
    + 2 (C \grad f) \cdot \grad g
    + f \DD g ,
\end{align*}
where $\vec{c}_j = \sum_i \partial_j C_{ij}$.
So, with $f = \gamma \varphi_s$,
\begin{eqnarray*}
\Lgen_s g
    &=&
    \frac{r}{\varphi_s} \left[
        \DD^*(\gamma \varphi_s g) - \grad \cdot (\gamma \varphi_s g \meanq)
        - g \DD^*(\gamma \varphi_s) + g \grad \cdot (\gamma \varphi_s \meanq)
    \right] \\
    &=&
    \frac{r}{\varphi_s} \left[
        \gamma \varphi_s \DD g
        + 2 \gamma \varphi_s \vec{c} \cdot \grad g
        + 2 (C \grad (\gamma \varphi_s)) \cdot \grad g
        - \gamma \varphi_s \meanq \cdot \grad g
    \right] \\
    &=&
    r \gamma \left[
        \DD g
        + 2 \vec{c} \cdot \grad g
        + 2 (C \grad \log(\gamma \varphi_s)) \cdot \grad g
        - \meanq \cdot \grad g
    \right] ,
\end{eqnarray*}
which is equation \eqref{eqn:lineage_generator}.
